Combining living and fossil taxa into phylogenies: the missing data issue

### 2015-01-08 Manuscript/TEM_draft.tex
Discussion update

In this study, we focus on the effect of missing data on our ability to recover a "best" tree's topology because it is a crucial aspect of a phylogeny in many macroevolutionary studies, for example when trying to elucidate the evolutionary relationships among species \citep[e.g.][]{meredithimpacts2011,jetzthe2012}, or for studying evolutionary transitions \citep[e.g.][]{}. %CITE-check Lloyd and Friedman or birdy stuff
Although branch length estimation is also important (namely for timing extinction and/or speciation events - e.g. \citealt{ronquista2012}), we do not consider branch lengths in this study. This is partially due to difficulties with simulating branch lengths and topology simultaneously,
but also because previous studies have already empirically assessed the effect of the Total Evidence method on branch length variation but with a fixed topology approach \citep{ronquista2012,schragocombining2013,slaterphylogenetic2013,beckancient2014}.
Thus understanding the sensitivity of topology to missing data is important for assessing the accuracy of tree estimation in the Total Evidence framework. To our knowledge, this question has never been formally assessed.

Here we use a simulation approach to assess the effect of missing data on tree topologies inferred from Total Evidence matrices. Since the molecular part of a Total Evidence matrix acts like a "classical" molecular matrix containing only the living taxa \citep{ronquista2012}, the effect of missing data on such matrices is well known \citep{wiensmissing2006,wiensmissing2008,rouresite-specific2011}. Therefore, we focus only on missing data in the morphological part of the matrix. We investigate three major parameters that directly affect the completeness of the morphological part of the matrix: (i) the proportion of living taxa with no morphological data; (ii) the proportion of missing data in the fossil taxa; and (iii) the proportion of missing morphological characters for both living and fossil taxa in the matrix. We remove data from a Total Evidence matrix by changing the values of these three parameters and then assess how this affects the topology of trees inferred using Maximum Likelihood and Bayesian methods. We chose these parameters because they reflect empirical biases in data availability. The advent of molecular phylogenetics means that morphological data for living species is rarely collected, and few people have the skills to identify characters needed for detailed phylogenetic analysis. Missing data in fossil taxa is very common due to preservation biases \citep{sansomfossilization2013}, and the overall number of characters depends on the effort of the people identifying them \citep[e.g.][]{O'Leary08022013}. We find that the ability of Total Evidence methods to recover the "best" tree's topology is increased when using Bayesian methods. Also minimizing the proportion of living taxa with no morphological data and the proportion of missing morphological characters improves the ability of Total Evidence methods to recover the "best" tree's topology more so than minimizing the proportion of missing data in the fossil record.

@@

As the amount of missing data in the morphological part of the Total Evidence matrix increases, our ability to recover the "best" tree's topology decreases, regardless of the missing data parameter ($M_{L}$, $M_{F}$ or $M_{C}$), the tree inference method (Maximum Likelihood or Bayesian) or the tree comparison metric used (Normalised Robinson-Foulds or Normalised Triplets distance). However, the different missing data parameters and tree inference methods do not affect the topology in the same way (Fig.~\ref{Fig_Results-permeth_perparam} and Fig.~\ref{Fig_Results-global_perparam}).

\subsection{Individual effects of missing data parameters}
As the amount of missing data increases across all three parameters, our ability to recover the "best" tree's topology decreases (Fig.~\ref{Fig_Results-permeth_perparam}).
The Normalised Robinson-Foulds distance is always substantially lower for the Maximum Likelihood %(minimum Normalised Robinson-Foulds distance for $M_{L}$, $M_{F}$ and $M_{C}$ of respectively @, @ and @) %minimum RF value for each parameter individually% TG: or useless (just Bhattacharrya)?
 trees than for the Bayesian consensus trees %(minimum Normalised Robinson-Foulds distance for $M_{L}$, $M_{F}$ and $M_{C}$ of respectively @, @ and @)  %minimum Tr value for each parameter individually% TG: or useless (just Bhattacharrya)?
(median Bhattacharrya Coefficient = @, @, @ %BC value for each parameter individually between both methods RF
$M_{L}$, $M_{F}$ and $M_{C}$ respectively; Fig.~\ref{Fig_Results-permeth_perparam}; \hyperref[SupplementaryMaterial]{Supplementary Material Section 3}). The Normalised Triplets distance is similar between the Maximum Likelihood trees and the Bayesian consensus trees for the $M_{L}$ parameter (median Bhattacharrya Coefficient = @ %BC value for ML between both methods
; Fig.~\ref{Fig_Results-permeth_perparam}; \hyperref[SupplementaryMaterial]{Supplementary Material Section 3}), however, for the two other parameters, $M_{F}$ and $M_{C}$, the Normalised Triplets distance is substantially higher for the Maximum Likelihood trees than for the Bayesian consensus trees (median Bhattacharrya Coefficient = @, @  %BC value for each parameter individually between both methods Tr
$M_{F}$ and $M_{C}$ respectively; Fig.~\ref{Fig_Results-permeth_perparam}; \hyperref[SupplementaryMaterial]{Supplementary Material Section 3}).

@@

% NC: Quick comment. You find that Bayesian consensus trees are "best". BUT I think you need to be careful here with what you mean by "best". They are most likely to recover a more similar topology to the best tree than if you do the crazy tree comparison pairs. But that doesn't mean they're best, just that they perform best in your simulations. They are only best for that. If you want to advise people to use a method that best recovers topology as measured using RF or Triplets, then Bayesian consensus is best. BUT generally that is not what people want. Probably need to think some more about this.
% TG: a note about the "best" tree's topology regarding your comment just after the discussion header: should I replace "correct" by "best" since that's how we're calling the "best" tree anyway (all in quotes)

@@

Our results show that the ability to recover the "best" tree's topology in a Total Evidence framework decreases with increasing amount of missing data, regardless of how data were removed or the method of tree inference used. However, these factors affected topological recovery in different ways and to different extents. 
Decreasing the overall number of morphological characters in the matrix ($M_{C}$) and the amount of living taxa with morphological data ($M_{L}$) had worst effects on topological recovery (Fig.~\ref{Fig_Results-paircomp_within}). Additionally, using Bayesian consensus trees recovered the "best" tree's topology more consistently than using Maximum Likelihood trees or Bayesian posterior trees distributions (Fig.~\ref{Fig_Results-global_perparam} and Fig.~\ref{Fig_Results-paircomp_within}).
% ----->
% NC: Probably needs another summing up line here. Or maybe not
% <-----
% NC: Make sure it's parameters then tree building methods. NC: Make sure the order of the parameters is always the same L,F, C or whatever! THROUGHOUT the paper. Also correct RF and T to standardised OR normalised RF and T.
% NC: Might want to add a couple of subheadings to match those in the results.
As proposed in previous studies, our results show that the amount of missing data is not a problem \textit{per se} for Total Evidence methods, as long as enough morphological characters overlaps for most taxa in the matrix \citep[e.g.][]{kearneyfragmentary2002,wiensmissing2003,rouresite-specific2011,pattinsonphylogeny2014}. % NC: What do you mean by data overlap here? That species have overlapping characters?

\subsection{Effect of missing living taxa ($M_{L}$)} %Or remove? Also below then
%What is the parameter variation doing in our simulation
When the number of living taxa with morphological data ($M_{L}$) decreases, entire rows of data are being removed from the living taxa part of the matrix. Because living taxa still have 10 times more molecular characters available, even if they have no morphological data, the relation between them will always be fairly well resolved (depending on the phylogenetic signal from the molecular part of the matrix) and thus, ignoring the fossils, the living taxa clades will be always well conserved. However, this missing data parameter influences a lot the placement of fossil taxa because a decrease of the $M_{L}$ parameter will reduce the overlapping data between the living and the fossil taxa.

%Methods
%Part saying the difference between ML and Baycon
When using the Normalised Robinson-Foulds distance to measure topological similarity, the Bayesian consensus trees recover the "best" tree's topology more often than Maximum Likelihood trees (Fig.~\ref{Fig_Results-global_perparam}). In fact, regardless of the amount of missing data, Bayesian consensus trees always displays the highest Normalised Robinson-Foulds score (always $>$ 0.69; see \hyperref[SupplementaryMaterial]{Supplementary Material Section 3} for details). However, regarding the Normalised Triplets distance, the topological similarity with the "best" tree's topology decreases in the same way for both the Maximum Likelihood trees and the Bayesian consensus trees (Fig.~\ref{Fig_Results-permeth_perparam} and Fig.~\ref{Fig_Results-global_perparam}).

% ----->
% TG: Put that part somewhere in the start with the discussion of ML vs Baycon in the effect of ML part?
%-Probabilistic method always better (wright but see spencer) - Check McGuire[nope]/Sean[nope] literature for Bayesian vs. ML?
Variation in our ability to recover the "best" tree's topology also depended heavily on the method used to infer the tree (Fig.~\ref{Fig_Results-permeth_perparam} and Fig.~\ref{Fig_Results-global_perparam}). Previous studies have shown some superiority of methods using simple probabilistic models such as the M\textit{k} model \citep{lewisa2001} over cladistic methods (\citealt{wrightbayesian2014}; but see \citealt{spencerefficacy2013}). However, this is the first study, to our knowledge, that compares the performance of the M\textit{k} model \citep{lewisa2001} using Maximum Likelihood and Bayesian methods in a Total Evidence framework. %TG: I guess it's a good selling point: we ARE the first ones to do a sensitivity analysis in a TEM framework (as far as I know).
% <-----

%Part saying the difference between BS and baytrees
Regardless of the metric, the Maximum Likelihood boostrap trees and the Bayesian posterior trees distribution always performed poorlier in their ability to recover the "best" tree's topology, even when no morphological data was missing ($M_{L}$=0\% - Fig.~\ref{Fig_Results-permeth_perparam}). This reflects the fact it is difficult to compare two distributions of trees, and each comparison between a set of "missing data" tree and "best" tree in these cases involved 1000 random pairwise comparisons rather than just one. This will obviously add noise to the results for these two methods.

%Metrics
%Part explaining the differences between the metrics
This difference between the metrics is linked to what they measure and how they perform at doing it (see method %add link
, \hyperref[SupplementaryMaterial]{Supplementary Material S2} and \citealt{kuhnerpractical2014} for details on the nature of both metrics). In the Bayesian consensus tree topology, the clade conservation (i.e. Noramlised Robinson-Foulds distance) stays always high but the placement of wildcard taxa (i.e. the Normalised Triplets distance) decreases way faster (Fig.~\ref{Fig_Results-permeth_perparam} and Fig.~\ref{Fig_Results-global_perparam}). This is due to the fact that both metrics are illustrating two different aspects of tree topology (i.e. clade conservation for the Normalised Robinson-Foulds distance and wildcard taxa position for the Normalised Triplets distance - see method %add link!
and \hyperref[SupplementaryMaterial]{Supplementary Material S2} for details on both metrics). However, it is important to note that The Normalised Robinson-Foulds distance is more sensitive than the Normalised Triplets distance: when getting closer to the root of the tree, displacement of single taxon makes the clades not being exactly identical any more even if the clade still contains all the other taxa in both trees. Also, \citet{kuhnerpractical2014} have recently demonstrated that the Robinson-Foulds distance outperforms the Triplets distance for measuring topological differences between trees. Thus, the Bayesian consensus tree's topology seems to be the less affected by removing morphological data for living taxa (Fig.~\ref{Fig_Results-permeth_perparam} and Fig.~\ref{Fig_Results-global_perparam}).

\subsection{Effect of missing data for fossil taxa ($M_{F}$)}
When the overall proportion of data for the fossil taxa ($M_{F}$) decreases, this reduces the probability of morphological characters for the fossil taxa to overlap with the ones for the living taxa and leads to displacement of certain taxa. However, it is important to note that even though the number of displaced wildcard taxa increases (i.e. decrease of Normalised Triplets distance), the conservation of the clades (i.e. Normalised Robinson-Foulds distance) is still relatively good (mode=% value
%) when the proportion of missing data is high ($M_{F}$=75\%).

%Methods
Decrease in clade conservation is smaller than for param ML in the Bayesian consensus tree for the clade conservation but higher than param ML for the displacement of wildcard taxa. This is due to the fact that the Bayesian consensus tree is a majority consensus rule and that when the fossils have less data, they kind of randomly branch all over the shop and therefore no majority consensus position (>50\% of the time) exists puting the fossil on a politomie at the base of a clade (clade conservation stays high but displacement of taxa increases). On the opposite in MaxLik, the topology is always dichotomous but the fossil can be missplaced in the wrong clade (with a low BS value) but is moving less. In other words, Bayesian consensus tree places the fossil with a higher confidence at a low taxonomic level (politomy) and the MaxLik places the fossil with a low confidence at a high taxomomic level (baldy supported dichotomies). We can argue that the first option (higher confidence at a low taxonomic level) is preferable (example the Darwinius debate).

\subsection{Effect of missing character ($M_{C}$)}
When the overall number of morphological characters in the matrix ($M_{C}$) decreases
Story for MC: MC is expected to have an effect two times bigger than ML or MF, e.g. removing 10\% of MF removes 5\% of data (because it works only on the fossil part) and removing 10\% of MC actually removes 10\% of that.
%Method
However the effect of removing characters has an effect of topology of the same order of magnitude than for ML and MF (FigRes 1): For RF, Bayesian consensus is better than ML and for Tr, MaxLik is better than Baycon and have the same drop at 50\% as for MF (different than for ML). 
In this scenario, the overall number of missing characters ($M_{C}$) affects the proportion of overlapping data by reducing the overall size of the matrix (Fig.~\ref{Fig_Results-global_perparam}). % NC: What's the relevance of that sentence?
However, the pairwise Bhattacharyya Coefficients show that the area with the lowest probability of distributional overlap is led by comparing the trees with no missing data for the living taxa to the trees with 75\% missing data for the living data (right lower corner in Fig.~\ref{Fig_Results-paircomp_within}). %NC: I don't understand what you mean here

\subsection{Combined effect of missing data parameters ($M_{L}$, $M_{F}$ and $M_{C}$)} 
In our simulations, the overlap in data is greatly reduced when all three missing data parameters have a lot of missing data (e.g. $M_{L}$ = 75\%, $M_{F}$ = 75\% and $M_{C}$ = 75\%; Fig.~\ref{Fig_Results-permeth_perparam} and Fig.~\ref{Fig_Results-global_perparam}).
When combining the effect, we

%Our solutions
Our missing data parameters illustrate different sources of missing data in empirical matrices as follows: ($M_{L}$) the paucity of coded morphological characters for living taxa; ($M_{F}$) missing data for fossils (or parts of fossils) that have not been preserved in the fossil record; and ($M_{C}$) characters that have not been coded across living and fossil species, perhaps due to difficulties in coding or poor preservation of the feature in collections. Filling these gaps in empirical Total Evidence matrices should lead to a substantial increase in our ability to recover the "best" tree's topology. We can increase the number of living taxa with coded morphological characters by increasing research efforts in this area, and encouraging use of our vast natural history collections. Increasing data for fossil species is harder, since it depends on fossil preservation biases and new fossil discoveries. However, gaps in the matrix can be filled with future discoveries of exceptionally preserved fossils \citep[e.g.][]{nithe2013}. Fortunately, although this data is the most difficult to collect, it also has the least influence on whether our simulations recover the "best" tree's topology (Fig.~\ref{Fig_Results-paircomp_within}). Finally, although increasing the number of coded characters is relatively straightforward, the amount of time it takes to build a morphological matrix increases directly with the number of characters involved. One solution to this problem may be to engage with collaborative data collection projects through web portals such as \textit{morphobank} \citep{morphobank}, so that no one individual collects all the data.

%Opening/further work part - NC(meeting): Be less defensive (i.e. nothing seemed to have gone wrong THIS time).
%   -Simulating morpho characters
The variations in our results could also be explained by another methodological caveat: 
% NC: What variations do you mean? Only an issue if it might bias results in one direction or another
the performance of our simulation protocol. In order to reduce the computational time of our analysis ($~$150 CPU years), we ran our sensitivity analysis on modestly sized matrices of 1000 molecular characters and 100 morphological characters. These matrices are two orders of magnitude smaller than some matrices used in published analyses \citep[e.g.][]{springermacroevolutionary2012,nithe2013} but still ranges in the sizes of smaller matrices used in other ones \citep[e.g.][]{sallam2011craniodental,kellymolecular2014}. %TG: note about NOT citing O'Leary here. 1-I don't like the paper ; 2-the Ni et al paper (primates) as a big morpho matrix too (1k  characters against 4k for O'Leary, same order of magnitude?) ; 3-We might want to limit the papers we're citing? note about citing Sean's and your paper here: 1-is it okay (regarding last time warning of being extra careful when citing you're own work)?
Also, simulating morphological characters is more complex than molecular ones. The underlying pattern of their evolution is often more complex and ruled by more parameters than molecular characters (i.e the number of character states, the states frequencies, the substitution matrix and the statistical model used) \citep{Pagel22011994,wagner2000,lewisa2001,wrightbayesian2014}. Morphological characters studies involve many potential statistical pitfalls (e.g. independent characters violation, rate variation - \citet{davalosintegrating2014}) and especially (i) incongruence with molecular signal and (ii) homoplasy \citep{wagner2000}.
\begin{enumerate}
\item{First, morphological data can display a different signal than molecular data, especially in small matrices \citep{wagner2000}. %REF
This might lead to a controversial phylogenetic signal in the overall matrix and lower down the support values. However, regarding empirical data studies, most of the groups shows fairly congruent morphological and molecular phylogenetic signal \citep[e.g.][]{leerates2013}.}
\item{Secondly, in this study, we made the assumption that theoretically, morphological characters are randomly distributed on an organism however it seems clear that empirical morphological data does not act randomly \citep{sansomfossilization2013,pattinsonphylogeny2014}. However, following our simulation assumption of random character distributions, if they accumulate through time in the same way as the majority of the molecular characters then, homoplasic characters are expected to appear randomly through time \citep{davalosintegrating2014}. Therefore, homoplasy is expected to be more important (by chance) in bigger morphological matrices \citep{davalosintegrating2014}. After reaching a critical amount of morphological characters, adding new ones increases homoplasy \citep{wagner2000}.}
\end{enumerate}

>>>>

In this study, we focus on the effect of missing data on our ability to recover a "best" tree topology because it is a crucial aspect of a phylogeny in many macroevolutionary studies, for example when trying to elucidate the evolutionary relationships among species \citep[e.g.][]{meredithimpacts2011,jetzthe2012}, or for studying evolutionary transitions \citep[e.g.][]{}. %CITE-check Lloyd and Friedman or birdy stuff
Although branch length estimation is also important (namely for timing extinction and/or speciation events - e.g. \citealt{ronquista2012}), we do not consider branch lengths in this study. This is partially due to difficulties with simulating branch lengths and topology simultaneously,
but also because previous studies have already empirically assessed the effect of the Total Evidence method on branch length variation but with a fixed topology approach \citep{ronquista2012,schragocombining2013,slaterphylogenetic2013,beckancient2014}.
Thus understanding the sensitivity of topology to missing data is important for assessing the accuracy of tree estimation in the Total Evidence framework. To our knowledge, this question has never been formally assessed.

Here we use a simulation approach to assess the effect of missing data on tree topologies inferred from Total Evidence matrices. Since the molecular part of a Total Evidence matrix acts like a "classical" molecular matrix containing only the living taxa \citep{ronquista2012}, the effect of missing data on such matrices is well known \citep{wiensmissing2006,wiensmissing2008,rouresite-specific2011}. Therefore, we focus only on missing data in the morphological part of the matrix. We investigate three major parameters that directly affect the completeness of the morphological part of the matrix: (i) the proportion of living taxa with no morphological data; (ii) the proportion of missing data in the fossil taxa; and (iii) the proportion of missing morphological characters for both living and fossil taxa in the matrix. We remove data from a Total Evidence matrix by changing the values of these three parameters and then assess how this affects the topology of trees inferred using Maximum Likelihood and Bayesian methods. We chose these parameters because they reflect empirical biases in data availability. The advent of molecular phylogenetics means that morphological data for living species is rarely collected, and few people have the skills to identify characters needed for detailed phylogenetic analysis. Missing data in fossil taxa is very common due to preservation biases \citep{sansomfossilization2013}, and the overall number of characters depends on the effort of the people identifying them \citep[e.g.][]{O'Leary08022013}. We find that the ability of Total Evidence methods to recover the "best" tree topology is increased when using Bayesian methods. Also minimizing the proportion of living taxa with no morphological data and the proportion of missing morphological characters improves the ability of Total Evidence methods to recover the "best" tree topology more so than minimizing the proportion of missing data in the fossil record.

@@

As the amount of missing data in the morphological part of the Total Evidence matrix increases, our ability to recover the "best" tree topology decreases, regardless of the missing data parameter ($M_{L}$, $M_{F}$ or $M_{C}$), the tree inference method (Maximum Likelihood or Bayesian) or the tree comparison metric used (Normalised Robinson-Foulds or Normalised Triplets distance). However, the different missing data parameters and tree inference methods do not affect the topology in the same way (Fig.~\ref{Fig_Results-permeth_perparam} and Fig.~\ref{Fig_Results-global_perparam}).

\subsection{Individual effects of missing data parameters}
As the amount of missing data increases across all three parameters, our ability to recover the "best" tree topology decreases (Fig.~\ref{Fig_Results-permeth_perparam}).
The Normalised Robinson-Foulds distance is always substantially lower for the Maximum Likelihood %(minimum Normalised Robinson-Foulds distance for $M_{L}$, $M_{F}$ and $M_{C}$ of respectively @, @ and @) %minimum RF value for each parameter individually% TG: or useless (just Bhattacharrya)?
 trees than for the Bayesian consensus trees %(minimum Normalised Robinson-Foulds distance for $M_{L}$, $M_{F}$ and $M_{C}$ of respectively @, @ and @)  %minimum Tr value for each parameter individually% TG: or useless (just Bhattacharrya)?
(median Bhattacharrya Coefficient = @, @, @ %BC value for each parameter individually between both methods RF
$M_{L}$, $M_{F}$ and $M_{C}$ respectively; Fig.~\ref{Fig_Results-permeth_perparam}; \hyperref[SupplementaryMaterial]{Supplementary Material Section 3}). The Normalised Triplets distance is similar between the Maximum Likelihood trees and the Bayesian consensus trees for the $M_{L}$ parameter (median Bhattacharrya Coefficient = @ %BC value for ML between both methods
; Fig.~\ref{Fig_Results-permeth_perparam}; \hyperref[SupplementaryMaterial]{Supplementary Material Section 3}), however, for the two other parameters, $M_{F}$ and $M_{C}$, the Normalised Triplets distance is substantially higher for the Maximum Likelihood trees than for the Bayesian consensus trees (median Bhattacharrya Coefficient = @, @  %BC value for each parameter individually between both methods Tr
$M_{F}$ and $M_{C}$ respectively; Fig.~\ref{Fig_Results-permeth_perparam}; \hyperref[SupplementaryMaterial]{Supplementary Material Section 3}).

@@

% NC: Quick comment. You find that Bayesian consensus trees are "best". BUT I think you need to be careful here with what you mean by "best". They are most likely to recover a more similar topology to the best tree than if you do the crazy tree comparison pairs. But that doesn't mean they're best, just that they perform best in your simulations. They are only best for that. If you want to advise people to use a method that best recovers topology as measured using RF or Triplets, then Bayesian consensus is best. BUT generally that is not what people want. Probably need to think some more about this.
% TG: a note about the "best" tree topology regarding your comment just after the discussion header: should I replace "correct" by "best" since that's how we're calling the "best" tree anyway (all in quotes)

@@

Our results show that the ability to recover the "best" tree topology in a Total Evidence framework decreases with increasing amount of missing data, regardless of how data were removed or the method of tree inference used. However, these factors affected topological recovery in different ways and to different extents. 
Decreasing the amount of living taxa with morphological data ($M_{L}$) and the overall number of morphological characters in the matrix ($M_{C}$) had worst effects on topological recovery (Fig.~\ref{Fig_Results-paircomp_within}). Additionally, using Bayesian consensus trees recovered the "best" tree topology more consistently than using Maximum Likelihood trees or Bayesian posterior tree distributions (Fig.~\ref{Fig_Results-global_perparam} and Fig.~\ref{Fig_Results-paircomp_within}).As proposed in previous studies, our results show that the amount of missing data is not a problem \textit{per se} for Total Evidence methods, as long as enough morphological characters overlaps for most living and fossil taxa in the matrix \citep[e.g.][]{kearneyfragmentary2002,wiensmissing2003,rouresite-specific2011,pattinsonphylogeny2014}. % NC: What do you mean by data overlap here? That species have overlapping characters?

\subsection{Individual effects of missing data parameters}
\subsubsection{Effect of missing data for living taxa}
When the number of living taxa with morphological data ($M_{L}$) decreases, entire rows of data are being removed from the living taxa part of the matrix. Because living taxa still have 10 times more molecular characters available (see methods) %link to methods
, even if they have no morphological data, the relation between them will always be fairly well resolved (depending on the phylogenetic signal from the molecular part of the matrix) and thus, ignoring the fossils, the living taxa clades will be always well conserved. However, this missing data parameter influences a lot the placement of fossil taxa because a decrease of the $M_{L}$ parameter will reduce the overlapping data between the living and the fossil taxa.

When using the Normalised Robinson-Foulds distance to measure topological similarity, the Bayesian consensus trees recover the "best" tree topology more often than Maximum Likelihood trees (Fig.~\ref{Fig_Results-global_perparam}). In fact, regardless of the amount of missing data, Bayesian consensus trees always displays the highest Normalised Robinson-Foulds score (always $>$ 0.69; see \hyperref[SupplementaryMaterial]{Supplementary Material Section 3} for all the values). However, regarding the Normalised Triplets distance, the similarity with the "best" tree topology decreases in the same way for both the Maximum Likelihood trees and the Bayesian consensus trees for this particular missing data parameter (Fig.~\ref{Fig_Results-permeth_perparam} and Fig.~\ref{Fig_Results-global_perparam}).
% TG: Or is that line just repeating the results?
However, regardless of the metric, the Maximum Likelihood bootstrap trees and the Bayesian posterior trees distribution always performed poorlier in their ability to recover the "best" tree topology, even when no morphological data was missing ($M_{L}$ = 0\% - Fig.~\ref{Fig_Results-permeth_perparam}). This reflects the fact that it is difficult to compare two distributions of trees, and each comparison between a set of "missing data" trees and a set of the "best" trees involved 1000 random pairwise comparisons rather than just one. This will obviously add noise to the results for this method of measuring the ability to recover the "best" tree topology.

Variation in our ability to recover the "best" tree topology depends heavily on the tree inference method (Fig.~\ref{Fig_Results-permeth_perparam} and Fig.~\ref{Fig_Results-global_perparam}). For morphological data, previous studies have shown some superiority of probabilistic tree inference methods with simple evolutionary models such as the M\textit{k} model \citep{lewisa2001} over cladistic methods (\citealt{wrightbayesian2014}; but see \citealt{spencerefficacy2013}). However, this is the first study, to our knowledge, that compares the performance of the M\textit{k} model \citep{lewisa2001} for recovering the "best" tree topology using Maximum Likelihood and Bayesian methods in a Total Evidence framework. Our results shows that the topology of the Bayesian consensus tree is always closer to the "best" tree topology than the one of the Maximum Likelihood tree (Fig.~\ref{Fig_Results-global_perparam}).

In the Bayesian consensus tree topology, the clade conservation (i.e. approximated by the Normalised Robinson-Foulds distance) stays always high but the placement of wildcard taxa (i.e. approximated by the Normalised Triplets distance) decreases way faster (Fig.~\ref{Fig_Results-permeth_perparam} and Fig.~\ref{Fig_Results-global_perparam}). The difference between the metrics is linked to the way each is calculated and to how they perform at underlining changes in topology (see method %add link
, \hyperref[SupplementaryMaterial]{Supplementary Material S2} and \citealt{kuhnerpractical2014} for details on the nature of both metrics). In general, the Normalised Robinson-Foulds distance is more sensitive than the Normalised Triplets distance.
%----> version one
When getting closer to the root of the tree, the displacement of a single taxon will have a greater effect than it would in higher clades and therefore decrease the Normalised Robinson-Foulds distance.
%----> version two
%When getting closer to the root of the tree, displacement of single taxon makes the clades not being exactly identical any more even if the clade still contains all the other taxa in both trees.
Also, \citet{kuhnerpractical2014} have demonstrated that the Robinson-Foulds distance outperforms the Triplets distance for measuring topological differences between trees. Thus, the Bayesian consensus tree topology seems to be the less affected by removing morphological data for living taxa (Fig.~\ref{Fig_Results-permeth_perparam} and Fig.~\ref{Fig_Results-global_perparam}).

\subsubsection{Effect of missing data for fossil taxa}
When the overall proportion of data for the fossil taxa ($M_{F}$) decreases, this also reduces the probability of morphological characters for the fossil taxa to overlap with the ones for the living taxa. This can lead to difficulties for certain taxa to be branched in the tree (i.e. wildcard taxa behaviour). However, it is important to note that even though the number of displaced wildcard taxa increases (i.e. decrease of Normalised Triplets distance), the conservation of the clades (i.e. Normalised Robinson-Foulds distance) is still relatively good (mode = @% value
) when the proportion of missing data is high ($M_{F}$ = 75\%).

%----> version one
The effect of the missing data in the fossil record ($M_{F}$) on the Normalised Robinson-Foulds distance (i.e. as a proxy for clade conservation) is smaller than the one of the $M_{L}$ parameter in the Bayesian consensus trees (Fig.~\ref{Fig_Results-permeth_perparam} and Fig.~\ref{Fig_Results-global_perparam}). However, the effect of the $M_{F}$ parameter on the Normalised Triplets distance (i.e. as a proxy for displacement of wildcard taxa) is greater than the one of the $M_{L}$ parameter in the Bayesian consensus trees (Fig.~\ref{Fig_Results-permeth_perparam} and Fig.~\ref{Fig_Results-global_perparam}).
%----> version two
%The effect of the missing data in the fossil record ($M_{F}$) is smaller than the one the $M_{L}$ parameter for the clade conservation (Normalised Robinson-Foulds distance) but greater for the displacement of wildcard taxa (Normalised Triplets distance; Fig.~\ref{Fig_Results-permeth_perparam} and Fig.~\ref{Fig_Results-global_perparam}).
This can be related to the fact that the Bayesian consensus tree is built using a majority consensus rule. When a fossil taxa have less data (e.g. ($M_{F}$ = 75\%) they will tend to branch "randomly" to any taxa in the clade sharing the most characters. Therefore a majority consensus position is unlikely to exist (i.e. every branching position is represented in $<$ 50\% of the trees) and the fossil taxa will be branched on a polytomy at the base of the clade. This does not decreases the Normalised Robinson-Foulds distance (clades is conserved) but increases the Normalised Triplets distance (wildcard taxa is displaced). On the opposite, the same scenario in a Maximum Likelihood will lead to a dichotomous branching of the fossil but with a low bootstrap support ($<$ 50). In other words, the Bayesian consensus tree allows a fossil taxa with few data to be placed with a higher confidence at a lower taxonomic level than the Maximum Likelihood tree (where the fossil will be placed with lower confidence at a higher taxonomic position). We can argue that the first option (the Bayesian consensus tree topology) is preferable \cite[e.g][]{}. %Cite the Darwinius debate or is that a bit dangerous?

\subsubsection{Effect of missing character ($M_{C}$)}
The overall number of missing morphological characters is expected to have an effect on our ability to recover the "best" tree topology two times bigger than for the $M_{L}$ and $M_{F}$ parameters. For example, removing 10\% of the data for the fossil taxa actually removes a total of 5\% of data in the whole matrix (because this parameter affects only half of the taxa present in the matrix); on the other hand, removing 10\% of morphological characters actually removes 10\% of data in the matrix. However the effect of removing characters on the ability to recover the "best" tree topology is of the same order of magnitude than for the two other parameters (Fig.~\ref{Fig_Results-permeth_perparam}). Reducing the overall amount of morphological characters reduces the probability of overlap of morphological characters between the taxa present in the matrix and therefore decreases our ability to recover the "best" tree topology.

%I moved the stuff about simulating characters over here.
The overall number of morphological character in a Total Evidence matrix can also affect our ability to recover the "best" tree topology through two phylogenetic caveats from morphological data (in opposition to molecular data): (1) the incongruence of phylogenetic signal between morphological and molecular data and (2) homoplasy. The incongruence of phylogenetic signal between morphological and molecular data as previously been demonstrated as being potentially more important in small morphological matrices \citep{wagner2000}. Therefore, part of the decrease of the Normalised Robinson-Fould distance and the Normalised Triplets distance in our simulation could be due to conflicting phylogenetic signal between morphological and molecular data (Fig.~\ref{Fig_Results-permeth_perparam} and Fig.~\ref{Fig_Results-global_perparam}). However, regarding empirical data studies, most of the groups shows fairly congruent morphological and molecular phylogenetic signal \citep[e.g.][]{leerates2013,slaterphylogenetic2013,beckancient2014}. (2) Homoplasy, on the other hand is expected to increase with an increase of the morphological characters \citep{wagner2000}. %CITE
However, the use of probabilistic methods (i.e. Maximum Likelihood or Bayesian) and the M\textit{k} model \citep{lewisa2001} has been previously demonstrated to partially resolve this issue \citep{wrightbayesian2014}.

\subsection{Combined effect of missing data parameters}
As expected, when combining the missing data parameters, our ability to recover the "best" tree topology is affected in the same way as for the effect of the missing data parameters individually: the Normalised Robinson-Foulds distance and the Normalised Triplets distance is higher when all the missing data parameters have low missing data (i.e. $M_{L}$ = 0\%, $M_{F}$ = 0\%, $M_{C}$ = 0\%) and lower when they have high missing data (i.e. $M_{L}$ = 75\%, $M_{F}$ = 75\% and $M_{C}$ = 75\%; Fig.~\ref{Fig_Results-global_perparam}). However, it is important to notice that the effect of each parameter is not additive. The number of missing living taxa with morphological data ($M_{L}$) and the overall number of missing morphological character ($M_{C}$) have a bigger effect than the amount of missing data for the fossil taxa ($M_{F}$); when both $M_{L}$ and $M_{C}$ reach 50\% of missing data, the matrix does not contain enough phylogenetic information for the fossil taxa to be placed with confidence in the tree (Fig.~\ref{Fig_Results-paircomp_within}).

\subsection{Suggested applications} %Or "Practical implications"
Our missing data parameters illustrate different sources of missing data in empirical matrices as follows: ($M_{L}$) the paucity of coded morphological characters for living taxa; ($M_{F}$) the missing data for fossils (or parts of fossils) that have not been preserved in the fossil record; and ($M_{C}$) characters that have not been coded across living and fossil species, perhaps due to difficulties in coding or poor preservation of the feature in collections. Filling these gaps in empirical Total Evidence matrices should lead to a substantial increase in our ability to recover the "best" tree topology. We can increase the number of living taxa with coded morphological characters by increasing research efforts in this area, and encouraging use of our vast natural history collections. Increasing data for fossil species is harder, since it depends on fossil preservation biases and new fossil discoveries. However, gaps in the matrix can be filled with efforts in palaeontological field work that can potentially lead to future discoveries of exceptionally preserved fossils \citep[e.g.][]{nithe2013}. % TG: I added the palaeo field work just as a kind work to the pokemon masters that could read this paper - maybe useless?
Fortunately, although this data is the most difficult to collect, it also has the least influence on whether our simulations recover the "best" tree topology (Fig.~\ref{Fig_Results-paircomp_within}). Finally, although increasing the number of coded characters is relatively straightforward, the amount of time it takes to build a morphological matrix increases directly with the number of characters involved. One solution to this problem may be to engage with collaborative data collection projects through web portals such as \textit{morphobank} \citep{morphobank}, so that no one individual collects all the data.
